Demo: Counterpoint Analysis and Synthesis

### 2022-08-18 doc/farm22/abstract.tex
revised FARM paper

We describe at a high level some of the key features of the
library. Details can be found in the code repository
\citep{HaskellCounterpoint}.

At a high level music is represented as a list of voices, which in
turn are a list of bars, which are in turn a list of beats. Here
dependent types would be useful to ensure for example that every voice
has the same number of bars, but a run-time check can also be used to
enforce this. Each of these elements is parameterized, allowing a
number of pitch-like elements to be used to construct music. These
could be algebraic data types representing name (like G\#) and
octave, or a numeric representation of pitch. Either datum can then be
embedded into a wrapper, such as \texttt{MPitch} which can either be a
fixed pitch or a named hole to be solved for.

The above representation of music allows natural handling of first,
second and third species counterpoint (one to four notes per cantus
firmus note) as well as multi-part counterpoint. At a lower level the
music is translated into a simpler form---for example first-species
two part counterpoint is most simply represented as a list of pairs.

We plan to try to implement a constraint language as an embedded
DSL. However for now Haskell typeclasses are used, which have the
advantage that constraints can be written in natural Haskell; the only
drawback is that type signatures are somewhat abstract. The main
challenge is that we want to write constraints only once and use them
for both analysis and synthesis. The former expects low-level integers
and booleans while the latter their symbolic counterparts. This is
handled by using typeclasses which can be instantiated with either.

Analysis and synthesis then both take the same music and list of
constraints. The former checks the constraints returning error
messages in the higher-level language used to express the music. The
latter generates (if possible) music by filling in the given holes
using the constraints and the SMT solver. MIDI is also generated,
which can then be opened and viewed with notation software such as
MuseScore.

>>>>

We describe at a high level some of the key features of the
library; it is under development so details may change. The
latest code is available at \citep{MusicTools}.

The core component is a simple constraint language (\texttt{Expr}) of
booleans and integers. All constraints must be expressed at a low level
in this language. Integers represent absolute pitches and there are
named variables for synthesized pitches. Integer functions include
addition, subtraction and modulo. Equalities and inequalities convert
from integers to booleans; a monomorphic \texttt{if\_then\_else\_} allows
the reverse.

At a higher level, constraints (\texttt{Constraint}) are represented as
simple algebraic data types. For example \texttt{contrary}
denotes the \texttt{MotionConstraint} that two pairs of pitches
(each representing an vertical interval) represent contrary
motion. Constraints are compiled into boolean expressions, and only the
constraint author needs to understand these details; the user can
employ the high-level constraints directly.

At a higher level still, counterpoint rules (\texttt{Counterpoint})
take music as input and generate a list of constraints. For analysis of
music (no unknown pitches), constraints can be compiled and evaluated;
those which evaluate to false fail to hold. This can be done
interactively within emacs, for example.

For synthesis, the code must currently be compiled and run. The Agda
constraint language is converted to Haskell and compiled into symbolic
integers and booleans using SBV (\texttt{Smt}). An SMT solver
(currently only Z3 is supported) is called to solve for the variable
pitches, if possible. The result is then output as MIDI
(\texttt{SmtInterface}) which can then be opened and viewed with
notation software such as MuseScore.
